Natural selection in the Health and Retirement Study

### 2023-09-27 hrs-selection.qmd
Tweak discussion

@fig-true-effects reports estimates for the true PGS as well as the "true chip-PGS" - the true best predictor of the phenotype using only variants recorded on the array chip, for phenotypes where selection effects were nominally significant and a measure of the phenotype was available in-sample and was significantly predicted by the PGS. It's important to understand that estimates for the true PGS are more uncertain than for the true chip-PGS. Ideally, the true chip-PGS differs only from the recorded PGS because of sampling error in calculating the underlying summary statistics.[^2] The true PGS would use many more genetic variants, perhaps including e.g. *de novo* mutations, and the relationship between these variants' effect on fertility and on the target phenotype may not be the same as for genetic variants captured on array chips.

[^2]: But even the chip-PGS might contain non-random noise from correlated environments.

The HRS contains multiple PGS for several phenotypes, typically with the $R^2$ increasing over time. Appendix @fig-check-correction runs a sanity check in which I predict the coefficient of later scores on fertility, using earlier scores and the above formula. Mostly, earlier PGS do a good job at predicting the effect of later PGS on fertility, with the later PGS's point estimate within the earlier PGS's 95% confidence interval. One exception is that the point estimate of the earlier PGS for general cognition, GENCOG, on the effect of the later PGS GCOG2 has the wrong sign.

As a rough rule of thumb, a 0.1 standard deviation change in a polygenic score over a generation might count as "serious": about 54% of the new generation will be below the parents' mean. Many upper confidence bounds meet that threshold, but lower bounds are often small or include zero. Given the uncertainties about the exact nature of "noise" in existing PGS, the estimates -- even for chip heritability -- mostly show the limits of our knowledge, and should be treated as best guesses only.

>>>>

@fig-true-effects reports estimates for phenotypes where selection effects were nominally significant and a measure of the phenotype was available in-sample and was significantly predicted by the PGS. As a rule of thumb, a 0.1 standard deviation change in a polygenic score over a generation might count as "serious": about 54% of the new generation will be below the parents' mean. Many upper confidence bounds meet that threshold, but lower bounds are often small or include zero. The confidence bounds capture uncertainty from sampling variation, but not other sources, including uncertainty about the true $h^2$ , limitations of the within-sample phenotypes, noise from correlated environments, and for twin-heritability, different relationships with fertility among variants not measured on the chip. Given all this, the estimates mostly show the limits of our knowledge, and should be treated as best guesses only.

The HRS contains multiple PGS for several phenotypes, typically with the $R^2$ increasing over time. Appendix @tbl-check-correction runs a sanity check in which I predict the coefficient of later scores on fertility, using earlier scores and the above formula. Mostly, earlier PGS do a good job at predicting the effect of later PGS on fertility, with the later PGS's point estimate within the earlier PGS's 95% confidence interval. One exception is that the point estimate of the earlier PGS for general cognition, GENCOG, on the effect of the later PGS GCOG2 has the wrong sign.
